PACKER: An Exemplar Model of Category Generation

### 2017-01-30 Manuscripts/cogsci-2017/paper.tex
6 pagesgit pull

The creative generation of new concepts is an intriguing yet understudied topic in cognitive science. In this paper, we present a novel exemplar-based approach to category generation: PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}). PACKER's core design assumptions are that (\textit{a}) categories are represented as exemplars in a multidimensional psychological space, (\textit{b}) generated items should be similar to exemplars of the same category, and (\textit{c}) generated categories should be dissimilar to existing categories. A behavioral study and formal simulations reveal broad support for PACKER's predictions.

\textbf{Keywords:} 
Categorization, exemplar models, category generation, creative cognition.
\end{abstract}

@@

Recent work has proposed and tested computational models of the processes underlying category generation. \citet{jern2013probabilistic} trained participants on a set of experimenter-defined categories composed of exemplars within an artificial three-dimensional domain (e.g., 2D shapes varying in size, hue, and saturation). After a short training phase, participants were asked to generate exemplars from a new category. Participants were provided with a set of scales to adjust the feature values of each generated stimulus, and were given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that generated categories possessed the same feature variance and correlations as the experimenter-defined categories in the domain. They evaluated different possible computational models of how people generate categories (such as an exemplar-based "copy-and-tweak" model) and found that a hierarchical Bayesian model (which generates categories by abstracting the distributional structure of the domain) provided the strongest account of the behavioral results.

@@

In the sections below, we formally describe the PACKER model and explore its predictions in a behavioral experiment. We compare its performance to a copy-and-tweak and hierarchical Bayesian model by examining their fits to aggregate results and individual differences.

@@

The PACKER model is an extension of the prominent Generalized Context Model of category learning (GCM; \citealp{nosofsky1984choice}). The model assumes that each category is represented by a collection of exemplars within in a $k$-dimensional psychological space, and that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories.

As in the GCM, the similarity between two examples, $s\left(x_i, x_j\right)$, is an inverse-exponential function of distance:
\begin{equation}
  s\left(x_i,x_j\right) = \exp \left\{ -c \sum_{k}{ \left| x_{ik} - x_{jk} \right|}w_k \right\}
  \label{eq:similarity}
\end{equation}
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and $\sum_k{w_k} = 1$), accounting for the relative importance of each dimension when calculating similarity, and $c$ ($c>0$) is a specificity parameter controlling the spread of exemplar generalization. For simplicity, most simulations will use uniform attention weights. We will use non-uniform attention weights in our discussion of individual differences.

To generate a new example, the model considers both the similarity to examples from contrast categories as well as similarity to examples in the target category. The aggregated similarity $a$ between generation candidate $y$ and the model's stored exemplars $x$ can be computed as $a(y, x) = \sum_j{f(x_j) s(y, x_j)}$, where $f(x_j)$ is a function specifying the extent to which each exemplar contributes to the target category. For brevity, we will abbreviate $f(x_j)$ as $f_j$. PACKER sets $f_j$ depending on exemplar $j$'s category membership: $f_j = \phi$ if $x_j$ is a member of the contrast category, and $f_j = \gamma$ if $x_j$ is a member of the target category. $\phi$ and $\gamma$ are free parameters ($-\infty \leq \phi, \gamma \leq \infty$) controlling the contribution of contrast- and target-category similarity, respectively. Larger absolute values for either parameter produce greater consideration of those exemplars, with values of 0 producing no effect. A negative value for $f_j$ produces a `repelling' effect (exemplars are less likely to be generated nearby $x_j$). Conversely, a positive value for $f_j$ produces a `pulling' effect (exemplars are more likely to be generated nearby $x_j$).

Recall that we proposed that categories are generated by balancing two constraints: that new categories should be different from existing categories, and exemplars belonging to the same category should be similar to one another. This is realized in PACKER when $\phi < 0$, and $\gamma > 0$. Negative values for $\phi$ encourage $y$ to be distant from the contrast category (as similarity to contrast category exemplars are subtracted during aggregation). Positive values for $\gamma$ encourage $y$ to be close to other exemplars of the target category. When $|\phi| = \gamma$, the "force" of repulsion from the contrast category is equal to the "force" of attraction to the target category.

@@

\subsection{Summary}
The proposed PACKER model suggests people generate categories by minimizing between-category similarity and maximizing within-category similarity. The underlying processes assumed by PACKER are highly similar to those in the GCM, with the only alteration being that PACKER aggregates positive- and negative-valued similarities, rather than only aggregating positive-valued similarities.

In later sections, we will explore the unique predictions yielded by these design principles. First, however, we contrast PACKER with other category generation models.

@@

Previous models of category generation focus on capturing the tendency for people to produce new categories that have similar distributional properties to existing categories. To the best of our knowledge, \citet{jern2013probabilistic} were the first to quantitatively evaluate computational models of generation. Based on their work, we describe two alternative models: a formalization of the \textit{copy-and-tweak} hypothesis \citep{ward2002role,ward1995s}, and the \textit{hierarchical sampling} proposed by \citet{jern2013probabilistic}.

\subsection{Copy-and-Tweak}
The \textit{copy-and-tweak} model proposes that generation is a two-part process: First, learners retrieve an observation of the target class from memory, and then they tweak it make something new. \citet{jern2013probabilistic} interpreted this proposal in terms of an exemplar model using the GCM \citep{nosofsky1984choice}. Formally, their model is equivalent to PACKER with $\{\phi = 0$, $\gamma = 1\}$ (see Figure \ref{fig:example-prob-spaces}). To allow for more flexible performance, and to explicitly formalize the `tweak' procedure, we slightly generalized their implementation.

In our copy-and-tweak model, all observations $z$ from the target class are equally likely to be retrieved as the source: $p(z) = 1 / n$, where $n$ is the number of stored examples in the target class. If $n=0$, learners are assumed to generate an item at random. After a source exemplar is retrieved, the similarity between generation candidates $y$ and the retrieved exemplar is computed as per Equation \ref{eq:similarity}. The model's goal is to generate an item that is similar to $z$, but does not exceed a given similarity threshold: The generated item should be similar, but not too similar to $z$. Formally, the probability that candidate exemplar $y$ will be generated based on a source exemplar $z$ is

\begin{equation}
    p(y|z)  = \dfrac
    { \exp \left\{\theta \cdot s(y,z) \right\} I\left(s(y,z) \leq \tau\right) }
    {\sum_i{\exp \left\{ \theta \cdot s(y_i,z) \right\} I\left(s(y_i,z) \leq \tau\right)}} 
\end{equation}
% 
where $\theta$ is a response determinism parameter, $\tau$ is the similarity threshold, and $I(\cdot)$ is the indicator function, which returns 1 when it is passed a true expression and 0 otherwise. So, $I\left(s(y,z) \leq \tau\right)$ is 1 when the candidate exemplar $y$ is far enough away from the source exemplar $z$. This ensures the source exemplar is tweaked "enough". When $\tau=1$, the threshold has no effect, and our model is equivalent to that of \citet{jern2013probabilistic}. To obtain predictions not depending on a given source example, the model's predictions can be aggregated over all possible sources: $p(y) = \sum_z{p(z)p(y|z)}$.

@@

Based on several results inconsistent with the copy-and-tweak account, \citet{jern2013probabilistic} advocated a hierarchical Bayesian model. Exemplars of each category were generated from a multivariate Gaussian distribution over the dimensions of stimulus space. The mean of each category was independent, but the covariance matrices (encoding feature variances and correlations) were generated from a common prior distribution. New categories are generated then by generating a new mean (uniform over stimulus space) and covariance matrix from the common prior distribution. Because the shared prior distribution's parameters were unobserved, a hierarchical Bayesian model uses information from the previous categories (their feature variances and correlations) to generate the covariance matrix of the new category.

We assume that each category's exemplars are distributed according to a multivariate normal distribution with parameters ($\mu, \Sigma$). We assume that each category's covariance matrix is inverse-Wishart distributed with parameters ($v$, $\kappa,$ and $\Sigma_D$).\footnote{Note that \citet{jern2013probabilistic}'s model is slightly different, as they used a non-conjugate model. Their model acts very similar to our version of it and receives comparable fits.} $\Sigma_D$ is the covariance matrix shared between categories. We assumed the shared covariance matrix $\Sigma_D$ was generated from a Wishart distribution (for conjugacy) with parameters $v_0$, $\kappa_0$, and $\Sigma_0$. We set $\nu_0 = \kappa_0 = 1$, and $\Sigma_0 = \rho {\bf I}$, where $\rho$ was a free parameter controlling the expected variance of dimensions (dimensions of the shared covariance matrix are expected to be uncorrelated) and ${\bf I}$ is the identity matrix.

@@

Generated exemplars are then assumed to be drawn from the multivariate normal distribution specified by $(\mu_{B}, \Sigma_{B})$. Thus, $p(y)$ is
\begin{equation}
  p(y) = \dfrac
    {\exp( \theta \cdot {\rm Normal}(y; \mu_{B}, \Sigma_{B}))}
    {\sum_i \exp( \theta \cdot {\rm Normal}(y_i; \mu_{B}, \Sigma_{B}))} 
\end{equation}
where $\theta$ is a response determinism parameter and ${\rm Normal}(y; \mu, \Sigma)$ denotes a multivariate Normal density evaluated at $y$. 
%JLA: There's something funky going on with the notation. I got confused by your noation, so hopefully you can update what i did to be consistent with the rest.

@@

Although our manipulation is minimal, the PACKER model is capable of predicting strong between-condition differences. PACKER proposes that the nature of the space not occupied by the Alpha category determines where members of the Beta category are likely to be generated. Thus, the lower areas of the stimulus space should be less frequently used for generation in the Bottom condition compared to the Middle (as these areas possess greater similarity to the Bottom Alpha category). Conversely, the upper areas of the stimulus space should be used for generation more frequently in the Bottom condition compared to Middle.

@@

Following the training phase, participants were asked to generate four examples belonging to another category called `Beta'. Participants were instructed that members of the Beta category could be quite similar or different depending on what they think makes the most sense for the category, but that they were not allowed to make the same example twice. As in \citet{jern2013probabilistic}, generation was completed using a sliding-scale interface. Participants were presented with two scales that controlled the dimensions of the generated example. An on-screen preview of the example updated whenever one of the features was changed. Participants could generate any example along an evenly-spaced 9x9 grid, except for any previously generated Beta exemplars. Neither the members of the Alpha category nor the previously generated Beta examples were visible during generation.

@@

Because the conditions differ only in their location along the y-axis (see Figure \ref{fig:middle-bottom-conditions}), we focus on how Beta exemplars are generated above and below the contrast category. Several sample Beta categories are depicted in Figure \ref{fig:beta.samples}. As is evident in Figure \ref{fig:beta.samples}, we observed broad individual differences in generation strategy: Whereas some participants generated all four Beta examples within a narrow y-axis range, others generated Beta examples along a wide range.

@@

To evaluate PACKER's other predictions, we computed the number of exemplars produced at different distances to the center of the Alpha category. These data, depicted in the left panel of Figure \ref{fig:distance.figs} reveal a strong preference for stimuli that are dissimilar to the members of the Alpha category: maximally distant items were by far the most frequently generated.

Finally, we computed for each participant the average distance between exemplars belonging to the same and opposite categories. These data (depicted in the right panel of Figure \ref{fig:distance.figs}) show that, as observed by \citet{ward1994structured}, most people generated Beta categories in which members are closer to one another than they are to members of the Alpha category (i.e., more between- than within-category distance). We did however, observe a notable subset of individuals with greater within-class distance. These individuals tended to adopt a `corners' approach, in which a Beta examples were placed almost exclusively in the corners of the space.

@@

Our results support PACKER's predictions: People tend to generate items that are dissimilar from the contrast category and similar to the target category (with some exceptions, see Figure \ref{fig:distance.figs}). By consequence, we observed considerable differences in generation between the Middle and Bottom conditions: participants in the Bottom condition were less likely to use the bottom row of the stimulus space for generation, and participants in the Middle condition were more likely to create categories spanning the entire y-axis (utilizing the top and bottom row of the space). This latter result is especially interesting as it conflicts with previous results: Qualitatively different types of categories are generated (with very different distributional information), depending on only the location of the Alphas in the domain.

The results described above are somewhat commonsense: they simply demonstrate that the location (rather than distributional structure) of existing categories imposes constraints on generation because people tend to generate examples in areas not occupied by existing categories. This principle, however, is not predicted by existing models of generation -- these models are instead designed to explain distributional correspondences between generated and existing categories.

@@

Our model-fitting results follow directly from the design of our experiment. Because the copy-and-tweak and hierarchical sampling models are not influenced by the location of contrast categories within the space, they do not capture the broad tendency for generated categories to be dissimilar to existing classes.

@@

\subsection{Individual Differences}
%In future work, we intend to explore more systematically the role of PACKER's parameters in different subtypes of performance. For example, whereas the tight cluster profile could be explained with a moderately negative $\phi$ and a strongly positive $\gamma$, the four corners profile could be simulated with a strongly negative $\phi$ and $\gamma\leq0$.
%While it is important that any model be able to explain different types of performance with the same parameter values, these radical differences in generation are difficult to account for under a single set of parameters.
Our behavioral results showed that, in general, generated categories consist of items that are dissimilar to members of opposite categories, and similar to members of the their own category. However, beyond this broad description of performance, we observed a great deal of individual differences in generation style. Manually inspecting the data reveals four typical patterns (see Figure \ref{fig:beta.samples}): `corners' categories with one Beta example in each corner of the space, tight clusters, `column'-like categories varying along the y-axis but not the x-axis, and `row'-like categories varying along the x-axis but not the y-axis. Here, we explore the role of PACKER's attention weights ($w_1$ and $w_2$) in producing row-like or column-like categories.

@@

For each stimulus in the domain, we computed the difference in range between the features (e.g., $range(size) - range(color)$) across every generated category in which that stimulus was a member. Taking the average of these range differences yields a gradient describing how, on average, categories were distributed for each stimulus. These data, depicted in Figure \ref{fig:range-diff-gradient}, reveal a highly systematic relationship between category structure and category location. Whereas column-like categories more often include stimuli to the left or right of the Alpha class, row-like categories appear above and below the Alpha class.  Thus, participants modify the distributional structure of new categories to the maximize distance from the contrast category.

To simulate this finding, we set the attention weight parameters in PACKER and copy-and-tweak separately per participant. The other free parameters were set as in Table \ref{table:model-fits}. While there exist methods to find the optimal attention weights for a given classification \citep[see][]{vanpaemel2012using}, for ease of computation we simply approximated the weights based on the inverse of each feature's range: $w_k \propto range(k)^{-1}$. Thus, the Alpha and Beta categories are assumed to be distinct along dimensions that the Betas do not vary on. To simulate the hierarchical sampling model we set the domain covariance prior, $\Sigma_0$, proportional to the range of each feature (not using the inverse). We then simulated 50 Beta categories with each participant's weighting scheme to obtain a sense of how the relative importance of each dimension affects what types of categories are generated and where they are generated.

@@

\section{Discussion}
The creative use of conceptual knowledge is a fascinating yet understudied topic in categorization. In this paper, we presented a novel exemplar-based approach to explaining category generation. The PACKER model proposes that categories are represented as a collection of exemplars stored in memory, and that members of generated categories should be similar to one another, yet dissimilar to members of opposing categories. Given that exemplar models can be viewed as Importance-Sampling approximations of Bayesian models \citep{shi10}, PACKER can be viewed as a rational process model and approximating the expected density of a new category based on a contrast category.

In a behavioral study and subsequent formal modeling, we found broad support for the PACKER model. Participants in our study more frequently generated items that are distant from members of contrast categories, and they tended to generate categories with more within-class than between-class similarity. Likewise, we found that the \textit{location} of contrast categories shapes generation by imposing constraints on the areas of space that remain available for a new category. These effects are somewhat commonsense in that they simply reveal that people tend to generate items that are different from what already exists. However, existing models \citep[see][]{jern2013probabilistic} assume little about the role of contrast in category generation (in these models, the location of generated categories is randomly chosen). Our formal simulations reveal that PACKER provides a strong account of the phenomena observed behaviorally.

The PACKER model is, in general, highly expressive in its performance. Under different parameter settings it is capable of generating tightly clustered or highly distributed categories, and adjusting the distribution of categories along each feature. Future work will focus on exploring how individual differences in generation can be explained using PACKER, and whether the canonical effects observed by \citet{ward1994structured} and \citet{jern2013probabilistic} can be explained through exemplar-similarity alone.

>>>>

Generating new concepts is an intriguing yet understudied topic in cognitive science. In this paper, we present a novel exemplar model of category generation: PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}). PACKER's core design assumptions are (1) categories are represented as exemplars in a multidimensional psychological space, (2) generated items should be similar to exemplars of the same category, and (3) generated categories should be dissimilar to existing categories. A behavioral study and formal simulations broadly support PACKER's predictions.

\textbf{Keywords:} 
Categorization, exemplar models, category generation, creative cognition, computational modeling.
\end{abstract}

@@

Recent work has proposed and tested formal models of category generation. \citet{jern2013probabilistic} trained participants on experimenter-defined categories composed of exemplars within an artificial three-dimensional domain. After a short training phase, participants were asked to generate exemplars from a new category. Participants were provided with a set of scales to adjust the feature values of each generated stimulus, and were given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that generated categories possessed the same feature variance and correlations as the experimenter-defined categories in the domain. They evaluated different possible computational models of how people generate categories (such as an exemplar-based "copy-and-tweak" model) and found that a hierarchical Bayesian model (which generates categories by abstracting the distributional structure of the domain) provided the strongest account of the behavioral results.

@@

In the sections below, we formally describe the PACKER model and explore its predictions in a behavioral experiment. We compare its performance to copy-and-tweak and hierarchical Bayesian models by examining their fits to aggregate results and individual differences.

@@

The PACKER model is an extension of the Generalized Context Model of category learning (GCM; \citealp{nosofsky1984choice}). It assumes that each category is encoded by a set of exemplars within a $k$-dimensional psychological space, and that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories.

As in the GCM, the similarity between two examples, $s\left(x_i, x_j\right)$, is an inverse-exponential function of distance:
\begin{equation}
  s\left(x_i,x_j\right) = \exp \left\{ -c \sum_{k}{ \left| x_{ik} - x_{jk} \right|}w_k \right\}
  \label{eq:similarity}
\end{equation}
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and $\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling the spread of exemplar generalization. For simplicity, simulations will use uniform attention weights, except in our discussion of individual differences.

To generate a new example, the model considers both the similarity to examples from contrast categories as well as the similarity to examples (if any exist) in the target category. The aggregated similarity $a$ between generation candidate $y$ and stored exemplars $x$ is $a(y, x) = \sum_j{f(x_j) s(y, x_j)}$, where $f(x_j)$ is a function specifying the extent to which each exemplar contributes to the target category. For brevity, we abbreviate $f(x_j)$ as $f_j$. PACKER sets $f_j$ depending on exemplar $j$'s category membership: $f_j = \phi$ if $x_j$ is a member of the contrast category, and $f_j = \gamma$ if $x_j$ is a member of the target category. $\phi$ and $\gamma$ are free parameters ($-\infty \leq \phi, \gamma \leq \infty$) controlling the contribution of contrast- and target-category similarity, respectively (see Fig \ref{fig:example-prob-spaces}). Larger absolute values result in greater consideration of those exemplars, with values of 0 eliminating their effect. A negative value for $f_j$ produces a `repelling' effect (exemplars are less likely to be generated nearby $x_j$). Conversely, a positive value for $f_j$ produces an `attracting' effect (exemplars are more likely to be generated nearby $x_j$).

Recall that we proposed that categories are generated by balancing two constraints: New categories should be different from existing categories, and exemplars belonging to the same category should be similar to one another. This is realized in PACKER when $\phi < 0$, and $\gamma > 0$. Negative values for $\phi$ encourage $y$ to be distant from the contrast category (as similarity to contrast category exemplars are subtracted during aggregation). Positive values for $\gamma$ encourage $y$ to be close to other exemplars of the target category. When $|\phi| = \gamma$, the repulsion effect from the contrast category is equal to the attraction effect to the target category.

@@

\subsection{Summary}
The proposed PACKER model suggests people generate categories by minimizing between-category similarity and maximizing within-category similarity. The underlying processes assumed by PACKER are highly similar to those in the GCM. The main difference is PACKER aggregates positive- and negative-valued similarities, rather than only aggregating positive-valued similarities. In later sections, we will explore the unique predictions yielded by these design principles. First, however, we contrast PACKER with other category generation models.

@@

Previous models of category generation focus on capturing the tendency for people to produce new categories that have similar distributional properties to existing categories. To the best of our knowledge, \citet{jern2013probabilistic} were the first to evaluate computational models of generation. Based on their work, we describe two alternative models: a formalization of the \textit{copy-and-tweak} hypothesis \citep{ward1995s}, and the \textit{hierarchical sampling} hypothesis \citep{jern2013probabilistic}.

\subsection{Copy-and-Tweak}
The \textit{copy-and-tweak} model proposes that generation is a two-part process: First, learners retrieve an observation of the target class from memory, and then they tweak it to make something new. \citet{jern2013probabilistic} interpreted this proposal in terms of an exemplar model using the GCM \citep{nosofsky1984choice}. Formally, their model is equivalent to PACKER with $\{\phi = 0$, $\gamma = 1\}$ (see Figure \ref{fig:example-prob-spaces}). To allow for more flexible performance, and to explicitly formalize the `tweak' procedure, we slightly generalized their implementation.

In our copy-and-tweak model, all observations $z$ from the target class are equally likely to be retrieved as the source. The first exemplar is generated uniformly at random over stimulus space. After a source exemplar is retrieved, the similarity between generation candidates $y$ and the retrieved exemplar is computed as per Equation \ref{eq:similarity}. The model's goal is to generate an item that is similar to $z$, but does not exceed a given similarity threshold: The generated item should be copied, but tweaked. Formally, the probability that candidate exemplar $y$ will be generated based on a source exemplar $z$ is

\begin{equation}
    p(y|z)  = \dfrac
    { \exp \left\{\theta \cdot s(y,z) \right\} I\left(s(y,z) \leq \tau\right) }
    {\sum_i{\exp \left\{ \theta \cdot s(y_i,z) \right\} I\left(s(y_i,z) \leq \tau\right)}} 
\end{equation}
% 
where $\theta$ is a response determinism parameter, $\tau$ is the similarity threshold, and $I(\cdot)$ is the indicator function, which returns 1 when it is passed a true expression and 0 otherwise. So, $I\left(s(y,z) \leq \tau\right)$ is 1 when the candidate exemplar $y$ is tweaked "enough". When $\tau=1$, the threshold has no effect, and our model is equivalent to that of \citet{jern2013probabilistic}. To obtain predictions not depending on a given source example, the model's predictions can be aggregated over all possible sources: $p(y) = \sum_z{p(z)p(y|z)}$.

@@

Based on several results inconsistent with the copy-and-tweak account, \citet{jern2013probabilistic} advocated a hierarchical Bayesian model. Exemplars of each category were generated from a multivariate Normal distribution over the dimensions of stimulus space. The mean of each category was independently generated, but the covariance matrix (encoding feature variances and correlations) was generated from a common prior distribution. New categories are produced by generating a new mean (uniform over stimulus space) and covariance matrix from the common prior distribution. Because the shared prior distribution's parameters were unobserved, a hierarchical Bayesian model uses information from the previous categories (their feature variances and correlations) to generate the covariance matrix of the new category.

Each category's exemplars are assumed to be a multivariate Normal distribution with parameters ($\mu, \Sigma$). Each category's covariance matrix is assumed to be inverse-Wishart distributed with parameters ($v$, $\kappa,$ and $\Sigma_D$).\footnote{Note that \citet{jern2013probabilistic}'s model is slightly different, as they used a non-conjugate model. Their model acts very similar to our version of it and receives comparable fits.} $\Sigma_D$ is the covariance matrix shared between categories. We assume the shared covariance matrix $\Sigma_D$ is generated from a Wishart distribution (for conjugacy) with parameters $v_0$, $\kappa_0$, and $\Sigma_0$. We set $\nu_0 = \kappa_0 = 1$, and $\Sigma_0 = \rho {\bf I}$, where $\rho$ is a free parameter controlling the expected variance of dimensions (dimensions of the shared covariance matrix are expected to be uncorrelated) and ${\bf I}$ is the identity matrix.

@@

Generated exemplars are drawn from a multivariate Normal distribution specified by $(\mu_{B}, \Sigma_{B})$. Thus, $p(y)$ is
\begin{equation}
  p(y) = \dfrac
    {\exp( \theta \cdot {\rm Normal}(y; \mu_{B}, \Sigma_{B}))}
    {\sum_i \exp( \theta \cdot {\rm Normal}(y_i; \mu_{B}, \Sigma_{B}))} 
\end{equation}
where $\theta$ is a response determinism parameter and ${\rm Normal}(y; \mu, \Sigma)$ denotes a multivariate Normal density evaluated at $y$. 
%JLA: There's something funky going on with the notation. I got confused by your noation, so hopefully you can update what i did to be consistent with the rest.

@@

Although our manipulation is minimal, the PACKER model predicts strong between-condition differences. PACKER proposes that the nature of the space not occupied by the Alpha category determines where members of the Beta category are likely to be generated. Thus, the lower areas of the stimulus space should be less frequently used for generation in the Bottom condition compared to the Middle (as these areas possess greater similarity to the Bottom Alpha category). Conversely, the upper areas of the stimulus space should be used for generation more frequently in the Bottom condition compared to Middle.

@@

Following the training phase, participants were asked to generate four examples belonging to another category called `Beta'. Participants were instructed that members of the Beta category could be quite similar or different depending on what they think makes the most sense for the category, but that they were not allowed to make the same example twice. As in \citet{jern2013probabilistic}, generation was completed using a sliding-scale interfaceTwo scales controlled the dimensions of the generated example. An on-screen preview of the example updated whenever one of the features was changed. Participants could generate any example along an evenly-spaced 9x9 grid, except for any previously generated Beta exemplars. Neither the members of the Alpha category nor the previously generated Beta examples were visible during generation.

@@

Because the conditions differ only in their location along the y-axis, we focus on how Beta exemplars are generated above and below the contrast category. Several sample Beta categories are depicted in Figure \ref{fig:beta.samples}. As is evident in Figure \ref{fig:beta.samples}, we observed broad individual differences in generation strategy: Whereas some participants generated all four Beta examples within a narrow y-axis range, others generated Beta examples along a wide range.

@@

To evaluate PACKER's other predictions, we computed the number of exemplars produced at different distances to the center of the Alpha category. These data (Figure \ref{fig:distance.figs} left) reveal a strong preference for stimuli that are dissimilar to the Alpha category members: maximally distant items were by far the most frequently generated.

Finally, we computed for each participant the average distance between exemplars belonging to the same and opposite categories. These data (Figure \ref{fig:distance.figs} right) show that, as observed by \citet{ward1994structured}, most people generated Beta categories in which members are closer to one another than they are to members of the Alpha category (i.e., more between- than within-category distance). We did however, observe a notable subset of individuals with greater within-class distance. These individuals tended to adopt a `corners' approach, in which a Beta examples were placed almost exclusively in the corners of the space.

@@

Our results support PACKER's predictions: People tend to generate items that are dissimilar from the contrast category and similar to the target category. We observed considerable differences in generation between the Middle and Bottom conditions: Participants in the Bottom condition were less likely to use the bottom row of the stimulus space for generation, and participants in the Middle condition were more likely to create categories spanning the entire y-axis (utilizing the top and bottom row of the space). This latter result is especially interesting as it conflicts with previous results: Qualitatively different types of categories were generated, depending only on the location of the Alphas.

Some aspects of the results described above are somewhat commonsense: They demonstrate that the location of existing categories also imposes constraints on generation because people tend to generate examples in areas not occupied by existing categories. This principle, however, is not predicted by existing models of generation -- these models were designed to explain distributional correspondences between generated and existing categories.

@@

Our model-fitting results make sense given the design of the experiment. As the copy-and-tweak and hierarchical sampling models are not influenced by the location of contrast categories within the space, they do not capture the broad tendency for generating categories dissimilar to existing classes.

@@

\subsection{Individual Differences}
%In future work, we intend to explore more systematically the role of PACKER's parameters in different subtypes of performance. For example, whereas the tight cluster profile could be explained with a moderately negative $\phi$ and a strongly positive $\gamma$, the four corners profile could be simulated with a strongly negative $\phi$ and $\gamma\leq0$.
%While it is important that any model be able to explain different types of performance with the same parameter values, these radical differences in generation are difficult to account for under a single set of parameters.
Generally, our behavioral results showed that generated categories consist of items that are dissimilar to members of opposite categories, and similar to members of their own category. We also observed a great deal of individual differences in generation style. Manually inspecting the data reveals four typical patterns (see Figure \ref{fig:beta.samples}): `corners' categories with one Beta example in each corner of the space, tight clusters, `column'-like categories, and `row'-like categories. Here, we explore the role of PACKER's attention weights ($w_1$ and $w_2$) in producing row-like or column-like categories.

@@

For each stimulus in the domain, we computed the difference in range between the features ($range(size) - range(color)$) across every generated category that had the stimulus as member. Taking the average of these range differences yields a gradient describing how, on average, categories were distributed for each stimulus. These data (Fig \ref{fig:range-diff-gradient}) reveal a systematic relationship between category structure and category location. Whereas column-like categories more often include stimuli to the left or right of the Alpha class, row-like categories appear above and below the Alpha class.  Thus, participants modify the distributional structure of new categories to the maximize distance from the contrast category.

To simulate this finding, we set the attention weight parameters in PACKER and copy-and-tweak per participant. The other free parameters were set as in Table \ref{table:model-fits}. While there exist methods to find the optimal attention weights for a given classification \citep[see][]{vanpaemel2012using}, for simplicity we approximated the weights using the inverse of each feature's range: $w_k \propto range(k)^{-1}$. Thus, the Alpha and Beta categories are assumed to be distinct along dimensions that the Betas do not vary on. To simulate the hierarchical sampling model we set the domain covariance prior, $\Sigma_0$, proportional to the range of each feature. We then simulated 50 Beta categories with each participant's weighting scheme to obtain a sense of how the relative importance of each dimension affects what types of categories are generated and where they are generated.

@@

\section{Discussion}
The creative use of conceptual knowledge is a fascinating yet understudied topic in categorization. In this paper, we presented a novel exemplar-based approach to explaining category generation. The PACKER model proposes that categories are represented as a collection of exemplars stored in memory, and that members of generated categories should be similar to one another, yet dissimilar to members of opposing categories. Exemplar models can be viewed as Importance-Sampling approximations of Bayesian models \citep{shi10}. So, PACKER can be viewed as a rational process model, approximating the expected density of a new category based on a contrast category.

In a behavioral study and subsequent formal modeling, we found broad support for the PACKER model. Participants in our study more frequently generated items that are distant from members of contrast categories, and they tended to generate categories with more within-class than between-class similarity. Likewise, we found that the \textit{location} of contrast categories shapes generation by imposing constraints on the areas of space that remain available for a new category. These effects reveal that people tend to generate items that are different from what already exists. However, existing models \citep[see][]{jern2013probabilistic} do not account for these effects. Our formal simulations reveal that PACKER provides a strong account of the phenomena observed behaviorally.

The PACKER model is, in general, highly expressive in its performance. Under different parameter settings it is capable of generating tightly clustered or highly distributed categories, and adjusting the distribution of categories along each feature. Future work will focus on exploring how individual differences in generation, and whether PACKER can explain previous results \citep{ward1994structured,jern2013probabilistic}.
